---
title: "Story 6 KDD"
author: "Keith DeNivo"
format: revealjs
editor: visual
---

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(ggplot2)
library(ggokabeito)
library(knitr)
library(RCurl)
library(kableExtra)
library(readr)
library(lubridate)
library(ggrepel)
library(DAAG)
library(FactoMineR)
library(broom)
library(tidymodels)
library(marginaleffects)
library(mlr3verse)
library(skimr)
library(patchwork)
library(factoextra)
library(cluster)
library(dbscan)
library(e1071)
library(fpc)
library(GGally)
library(kernlab)
library(mclust)
library(mlbench)
library(scatterpie)
library(seriation)
library(here)
library(ggdendro)
library(cowplot)
library(colorspace)
library(kableExtra)
library(ggforce)
library(ggiraph)
library(sf)
library(maps)
library(usmapdata)
library(usmap)
library(ggokabeito)
library(cowplot)
library(wordcloud)
library(tm)




```

```{r}
#| echo: false
#| include: false


userfeat_df <- data.frame(read.csv("C:/Users/keith/OneDrive - City University of New York/CUNY SPS MS DATA SCIENCE/Data 608/data 608 assignments/Story 6/Story - 6  Customer Segmentation at Instacart attached files May 20, 2025 914 PM/DataStory6_files/user_features.csv"))

depart_df <- data.frame(read.csv("C:/Users/keith/OneDrive - City University of New York/CUNY SPS MS DATA SCIENCE/Data 608/data 608 assignments/Story 6/Story - 6  Customer Segmentation at Instacart attached files May 20, 2025 914 PM/DataStory6_files/departments.csv"))

prod_df <- data.frame(read.csv("C:/Users/keith/OneDrive - City University of New York/CUNY SPS MS DATA SCIENCE/Data 608/data 608 assignments/Story 6/Story - 6  Customer Segmentation at Instacart attached files May 20, 2025 914 PM/DataStory6_files/products.csv"))
```

```{r}
#| echo: false
#| include: false



    
    
```

```{r}
#| echo: false
#| include: false


user <- as.data.frame( userfeat_df |> scale()) |> select(-user_id)

# pca_fit <-  user |> 
#   select(where(is.numeric))  |>  
#   prcomp(scale = TRUE)
```

```{r}
#| echo: false
#| include: false


# pca_fit |> 
#  augment(userfeat_df)  |> 
#   ggplot(aes(.fittedPC1, .fittedPC2)) + 
#   geom_point(size = 1.5) +
#   theme_minimal(base_size=16) +
#   labs(title="",
#        x = "PC 1", y= "PC 2",subtitle = "")
```

```{r}
#| echo: false
#| include: false


# set.seed(35)
# 
# wine_clusters = pca_fit|> 
#     augment(userfeat_df) |> 
#   select(.fittedPC1:.fittedPC13) |> 
#   kmeans(centers=3) |> augment(userfeat_df)
# 
# pca_fit |> augment(wine_clusters) |> 
#   ggplot(aes(.fittedPC1, .fittedPC2,color=.cluster)) + 
#   geom_point(size = 1.5) +
#   theme_minimal(base_size=16) +
#   labs(title="",
#        x = "PC 1", y= "PC 2",subtitle = "")
```

```{r}
#| echo: false
#| include: false


# set.seed(35)
# 
# wine_clusters = pca_fit|> 
#     augment(userfeat_df) |> 
#   select(.fittedPC1:.fittedPC13) |> 
#   kmeans(centers=3) |> augment(userfeat_df)
# 
# pca_fit |> augment(wine_clusters) |> 
#   ggplot(aes(.fittedPC1, .fittedPC2,color=.cluster)) + 
#   geom_point(size = 1.5) +
#   theme_minimal(base_size=16) +
#   labs(title="",
#        x = "PC 1", y= "PC 2",subtitle = "")
```

```{r}
#| echo: false
#| include: false

percent_df <- userfeat_df |> 
 mutate(row_total = rowSums(across((last_col()-6):last_col()))) |> 
 mutate(across(2:(ncol(userfeat_df) - 7), ~ .x / row_total, .names = "norm_{.col}")) |> 
select(starts_with("norm_"),user_id, Saturday, Sunday, Monday, Tuesday, Wednesday, Thursday, Friday)

head(userfeat_df)

```

```{r}
#| echo: false
#| include: false

set.seed(35)
pca_fit2 <-  percent_df |> select(-user_id) |> 
  select(where(is.numeric))  |>  
  prcomp(scale = TRUE)

 kmodel <- pca_fit2|> 
    augment(percent_df) |> 
  select(.fittedPC1:.fittedPC13) |> 
  kmeans(centers=3,nstart = 10) 

user_clusters = kmodel |> augment(percent_df)





```

## How Do We Categorize over 200,000 Users of Instacart?

### By their purchases of various products offered by Instacart

```{r}
#| echo: false



words <- colnames(userfeat_df)

words <- gsub("\\.", " ", colnames(userfeat_df))


wordcloud(words = words)
         


```

## Break down the purchase History by Principle Component Analysis

```{r}
#| echo: false


pca_fit2 |> augment(user_clusters) |> 
  ggplot(aes(.fittedPC1, .fittedPC2)) + 
  geom_point(size = 1.5) +
  theme_minimal(base_size=16) +
  labs(title="",
       x = "PC 1", y= "PC 2",subtitle = "")
```

## What are the main principle Components?

```{r}
#| echo: false

arrow_style <- arrow(
  angle = 20, length = grid::unit(8, "pt"),
  ends = "first", type = "closed"
)
pca_fit2 |> 
  tidy(matrix = "rotation")  |> 
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  )  |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text_repel(aes(label = column), hjust = 1,vjust=1) +
  xlim(-0.25, 0.5) + ylim(-0.25, 0.3) + 
  coord_fixed() +
  theme_minimal(base_size =16)+
  labs(title = "Contributing Purchases")
```

```{r}
#| echo: false
#| include: false


loadings <- as.data.frame(pca_fit2$rotation)


squared_loadings <- loadings^2


contribpca_df <- squared_loadings[, 1:5]
contribpca_df$total_contrib <- rowSums(contribpca_df)


contribpca_df <- contribpca_df[order(-contribpca_df$total_contrib), ]


contribpca_df <- contribpca_df |> 
 rownames_to_column(var = "variable")

head(contribpca_df)
```

```{r}
#| echo: false
#| include: false


#Dim 1
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC1`))  |> 
arrange(desc(total_contrib)) |> 
  slice_head(n=10)

#Dim 2
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC2`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)

#Dim 3
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC3`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)

```

## Which of the Purchase History Variables Separate Instacart Users?

```{r}
#| echo: false

#PC1 > PC2 for # Purchases on Days of the Week.

#PC2 > PC1  Goods Purchases

top_pc1 <- contribpca_df |> 
  group_by(variable) |> 
  summarise(contrib = sum(PC1)) |> 
  slice_max(contrib, n = 15) |> 
  mutate(PC = "PC1")

top_pc2 <- contribpca_df |> 
  group_by(variable) |> 
  summarise(contrib = sum(PC2)) |> 
  slice_max(contrib, n = 15) |> 
  mutate(PC = "PC2")




top_combined <- bind_rows(top_pc1, top_pc2,)


ggplot(top_combined, aes(x = reorder(variable, contrib), y = contrib, fill = PC)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Top Purchase History Variables for 
    PC1, PC2", subtitle = "PC1 > PC2 for # Purchases on Days of the Week
PC2 > PC1  Goods Purchases",
    x = "Product/Day",
    y = "Contribution"
  )

```

## Clustering Users via K-means  

```{r}
#| echo: false


kvec = seq(2:6)

wcss = map_dbl(kvec,.f = function(k) {
    kmeans(percent_df, centers = k, nstart = 5)$tot.withinss
})

elbow_plot = tibble(clusters = kvec, wcss = wcss) |> unnest()

elbow_plot |> ggplot(aes(kvec, wcss)) + 
  geom_line() +
  geom_vline(xintercept = 3, color = "red", linetype = 2) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(breaks=c(2,4,6,8,10)) +
  labs(title = "Elbow Plot",
       subtitle = "3 Clusters is optimal",
       y = "Within Cluster Sum of Squares",
       x = "Number of Clusters")
```

## Visualizing the Clusters

```{r}
#| echo: false


pca_fit2 |> augment(user_clusters) |> 
  ggplot(aes(.fittedPC1, .fittedPC2,color=.cluster)) + 
  geom_point(size = 1.5) +
  theme_minimal(base_size=16) +
  labs(title="3 Cluster of users",
       x = "PC 1", y= "PC 2",subtitle = "")
```

```{r}


```

```{r}
#| echo: false
#| include: false

 

  



percent_df2 = percent_df |> select(-user_id)





```

```{r}
#| echo: false
#| include: false

```

```{r}
#| echo: false
#| include: false



centers_mod <- kmodel$centers %*% t(pca_fit2$rotation[, 1:13])
centers_mod <- sweep(centers_mod, 2, pca_fit2$scale, "*")
centers_mod <- sweep(centers_mod, 2, pca_fit2$center, "+")


colnames(centers_mod) <- colnames(percent_df2)

centers_df <- as_tibble(centers_mod) |>
  mutate(center = factor(1:3))




```

## How Do the Clusters of Users Differ?

```{r}
#| echo: false

#Cluster 2 purchases the most often.

centers_df |> pivot_longer(cols = Saturday:Friday,names_to = "type") |> ggplot(aes(x=type,y=value)) + 
  geom_col(aes(fill = center)) +
  scale_fill_discrete(guide = "none") +
  geom_hline(yintercept = 0) +
  facet_wrap(~center,nrow=3) +
  theme_minimal(base_size = 12) +
  labs(title = "Clusters of Users Distinguished by Days", subtitle = "Cluster 2 purchases the most often.
Cluster 1 prefers Friday and Saturday Purchases
Cluster 3 purchases evenly throughout the week",
       y= "Scaled Values",
       x="Days")
```

```{r}
#| echo: false
#| include: false


```

```{r}
#| echo: false
#| include: false

```

## Users Prefer Fresh Fruits and Vegetables

```{r}
#| echo: false



#norm_fresh.vegetables , norm_paper.goods, norm_cleaning.products, norm_packaged.vegetables.fruits, norm_body.lotions.soap, norm_canned.jarred.vegetables, norm_oral.hygiene, norm_spices.seasonings, norm_fresh.vegetables, norm_dry.pasta, norm_dish.detergents, norm_packaged.produce


centers_df |> pivot_longer(cols = c(norm_fresh.vegetables , norm_paper.goods, norm_cleaning.products, norm_packaged.vegetables.fruits, norm_body.lotions.soap, norm_canned.jarred.vegetables, norm_oral.hygiene, norm_spices.seasonings, norm_dry.pasta, norm_dish.detergents, norm_packaged.produce, norm_fresh.herbs, norm_fresh.fruits, norm_laundry),names_to = "type") |> ggplot(aes(x=type,y=value)) + 
  geom_col(aes(fill=center)) +
  scale_fill_discrete(guide = "none") +
  geom_hline(yintercept = 0) +
  facet_wrap(~center,nrow=3) +
  theme_minimal(base_size = 10)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  
    labs(title = "Goods the different users purchase",
       y= "Scaled Values",
       x="Items")
```

## What about other Items?

```{r}
#| echo: false


centers_df |> pivot_longer(cols = c( norm_paper.goods, norm_cleaning.products, norm_packaged.vegetables.fruits, norm_body.lotions.soap, norm_canned.jarred.vegetables, norm_oral.hygiene, norm_spices.seasonings, norm_dry.pasta, norm_dish.detergents, norm_packaged.produce, norm_fresh.herbs, norm_laundry),names_to = "type")  |> ggplot(aes(x=type,y=value)) + 
  geom_col(aes(fill=center)) +
  scale_fill_discrete(guide = "none") +
  geom_hline(yintercept = 0) +
  facet_wrap(~center,nrow=3) +
  theme_minimal(base_size = 10)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  
    labs(title = "",subtitle = "cluster 3 focuses more on hygiene, paper, and cleaning products than other clusters
cluster 1 prefers packaged produce and generally avoids non food items
cluster 2 is a mix of cluster 3 and cluster 1",
       y= "Scaled Values",
       x="Items")
```
