---
title: "Do stricter gun laws reduce firearm gun deaths?"
author: "Keith DeNivo"
format: revealjs
editor: visual
---

```{r}
#| echo: false
#| include: false
library(readxl)
library(httr)
library(devtools)
library(RCurl)
library(plyr)
library(dplyr)
library(tidyverse)
library(DescTools)
library(ggpubr)
library(openintro)
library(readr)
library(rvest)
library(fpp3)
library(ggplot2)
library(tsibble)
library(feasts)
library(openxlsx)
library(latex2exp)
library(seasonal)
library(seasonalview)
library(fable)
library(rio)
library(urca)
library(zoo)
library(lubridate)
library(writexl)
library(janitor)
library(scales)
library(patchwork)
library(ggthemes)
library(ggrepel)
library(DT)
library(socviz)
library(sf)
library(maps)
library(geofacet)
library(statebins)
library(GGally)
```

```{r}
#| echo: false
#| include: false


c1 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%203/burglary%20(1).csv")
#read in the raw file
burglary_df <- data.frame(read.csv(text = c1 ))
#put the csv into a dataframe

c2 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%203/firearm_laws.csv")
#read in the raw file
laws_df <- data.frame(read.csv(text = c2 ))
#put the csv into a dataframe

c3 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%203/firearm_mortality2022.csv")
#read in the raw file
mortal_df <- data.frame(read.csv(text = c3 ))
#put the csv into a dataframe

c4 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%203/gun-ownership-by-state.csv")
#read in the raw file
own_df <- data.frame(read.csv(text = c4 ))
#put the csv into a dataframe

c5 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%203/mortality_breakdown.csv")
#read in the raw file
mbreak_df <- data.frame(read.csv(text = c5 ))
#put the csv into a dataframe
```

```{r}
#| echo: false
#| include: false


#convert 2 letter state abbreviations to the full state numaes
 mortal_df$STATE <- state.name[sapply(mortal_df$STATE, \(x) which(x == state.abb)[1])]
#add district of columbia
mortal_df$STATE[9] <- "District of Columbia"

mbreak_df$State <- state.name[sapply(mbreak_df$State, \(x) which(x == state.abb)[1])]

mbreak_df$State[9] <- "District of Columbia"

```

## The Amount of Gun Laws Vary Significantly by State

```{r}
#| echo: false


#graph the states with the least to most gun laws
laws_df |> as_tibble()  |> ggplot(aes(fill=LAWTOTAL,state=STATE)) + geom_statebins(lbl_size = 6) +
  scale_fill_gradient(low="white",high="blue") +
  theme_void(base_size=18) +
  labs(title="State Laws")




```

## Firearm Mortality Rate over time in New York

```{r}
#| echo: false


#graph the mortality rate change over time of new york
mortal_df |>  filter(STATE=="New York") |> 
  ggplot(aes(YEAR, RATE)) + 
  geom_area(fill = "#56B4E9", alpha = 0.7) +
  geom_line() + 
  scale_y_continuous(
    name = "firearm mortality rate",
    limits = c(0, 30), expand = c(0, 0),
   breaks = c(10, 20),
   labels = c( "10", "20")
  ) +
  scale_x_continuous(breaks = c(2010, 2015),
    labels = c("\'10", "\'15")) +
 coord_cartesian(clip = "off") +
 theme_minimal(base_size=20) +
  theme(
   axis.line.x = element_blank(),
    panel.spacing.x = grid::unit(1, "pt"),
    panel.spacing.y = grid::unit(1, "pt"),
   panel.grid.major = element_line(color = "gray80"),
   panel.background = element_rect(fill = "gray90")
  )+
  labs(title="New York Firearm Mortality Rate is
       Mostly Unchanged")
```

## Firearm Mortality either Increases Over Time or Remains the Same For Most States

```{r}
#| echo: false
#| warning: false


mortalmod_df <- mortal_df[-9, ]

#graph the mortality rate over time for all states
mortal_df |> 
  ggplot(aes(YEAR, RATE)) + 
  geom_area(fill = "#56B4E9", alpha = 0.7) +
  geom_line() + 
  scale_y_continuous(
    name = "firearm mortality rate",
    limits = c(0, 30), expand = c(0, 0),
    breaks = c(10, 20),
    labels = c( "10", "20")
  ) +
  scale_x_continuous(breaks = c(2010, 2015),
    labels = c("\'10", "\'15")) +
  coord_cartesian(clip = "off") +
  facet_geo(~STATE, grid = "us_state_grid1") +
  theme_minimal(base_size=11) +
  theme(
    axis.line.x = element_blank(),
    panel.spacing.x = grid::unit(1, "pt"),
    panel.spacing.y = grid::unit(1, "pt"),
    panel.grid.major = element_line(color = "gray80"),
    panel.background = element_rect(fill = "gray90")
  ) 
```

```{r}
#| echo: false
#| include: false


burglary_df <- burglary_df |> 
mutate(State = str_trim(State, side = "left"))
head(burglary_df)

burglary_df |> as_tibble()  |> ggplot(aes(fill=Burglary_100k,state=State)) + geom_statebins(lbl_size = 6) +
  scale_fill_gradient(low="white",high="red") +
  theme_void(base_size=18) +
  labs(title="State Burglary")
```

```{r}
#| echo: false
#| include: false


#more data cleaning


burglary_df <- burglary_df |> 
 rename(STATE = State)

mbreak_df <- mbreak_df |> 
 rename(STATE = State)

own_df <- own_df |> 
 rename(STATE = state)


burglary_df <- burglary_df |> 
 mutate(STATE = trimws(STATE))


```

```{r}
#| echo: false
#| include: false

data_frames <- list(burglary_df |> select(STATE,Burglary_100k), laws_df |> select(STATE, LAWTOTAL), mbreak_df , mortal_df |> filter(YEAR=="2022") |> select(STATE,RATE), own_df |> select(STATE,GunOwnershipPercentage))
merged_df <- Reduce(function(x, y) merge(x, y, by = "STATE", all = TRUE), data_frames)


```

```{r}
#| echo: false
#| include: false


#remove District of Columbia due to it giving issues.
merged_df <- merged_df[-9, ]

```

## What Are State Laws Correlated with?

#### Comparison with total gun laws, burglaries per 100k people, suicide rate, homicide rate, % gun ownership, and firearm mortality rate

```{r}
#| echo: false

ggpairs(merged_df[, 2:7])




```

## Gun Ownership and Mortality Rate

```{r}
#| echo: false


correlation <-  cor(merged_df$GunOwnershipPercentage, merged_df$RATE)
merged_df |> ggplot(aes(x= GunOwnershipPercentage, y=RATE))+
  geom_point(size=3)+
 geom_smooth(method = "lm", se = FALSE) +
annotate("text", x = Inf, y = Inf, label = paste("Correlation:", round(correlation, 3)),  hjust = 1, vjust = 2.1, size = 5, color = "blue") +
 labs(title = "The Firearm Mortality Rate 
  increases with Gun Ownership", x = "% Gun Ownership", y = "Mortality Rate")+
  theme_minimal(base_size = 20)
```

## Gun Ownership is Low in States with High Gun Laws

```{r}
#| echo: false


correlation <-  cor(merged_df$LAWTOTAL, merged_df$GunOwnershipPercentage)

merged_df |> ggplot(aes(x= LAWTOTAL, y=GunOwnershipPercentage))+
  geom_point(size = 3) +
 geom_smooth(method = "lm", se = FALSE) +
annotate("text", x = Inf, y = Inf, label = paste("Correlation:", round(correlation, 3)),  hjust = 1.1, vjust = 10, size = 5, color = "blue") +
 labs(title = "As the Total Gun Laws Increase
      the % of Gun Ownership Decreases", x = "Total Gun Laws", y = "% Gun Ownership")+
  theme_minimal(base_size = 20)

```

## Firearm Mortality Rate is Lower in States with High Gun Laws

```{r}
#| echo: false


correlation <-  cor(merged_df$LAWTOTAL, merged_df$RATE)
merged_df |> ggplot(aes(x= LAWTOTAL, y=RATE))+
  geom_point(size=3)+
 geom_smooth(method = "lm", se = FALSE) +
annotate("text", x = Inf, y = Inf, label = paste("Correlation:", round(correlation, 3)),  hjust = 1.1, vjust = 10, size = 5, color = "blue") +
 labs(title = "As the number of Gun laws Increase
      The Mortality Rate Decreases", x = "Total Gun Laws", y = "Mortality Rate")+
  theme_minimal(base_size = 20)
```
