---
title: "Data Story 5: Quantifying Factors that Control Housing Sales Prices"
author: "Keith DeNivo"
format: revealjs
editor: visual
---

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(ggplot2)
library(ggokabeito)
library(knitr)
library(RCurl)
library(kableExtra)
library(readr)
library(lubridate)
library(ggrepel)
library(DAAG)
library(FactoMineR)
library(broom)
library(tidymodels)
library(marginaleffects)
library(mlr3verse)
library(skimr)
library(patchwork)
library(factoextra)


```

## Quantifying Factors that Control Housing Sales Prices

### Ames Dataset

```{r}
#| echo: false


c1 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%205/train%20(1).csv")
#read in the raw file
train_df <- data.frame(read.csv(text = c1 ))
#put the csv into a dataframe

head(train_df)
```

```{r}
#| echo: false
#| include: false

#impute missing numeric values with median
numeric_cols <- sapply(train_df, is.numeric)
train_df[numeric_cols] <- lapply(train_df[numeric_cols], function(x) {
 x[is.na(x)] <- median(x, na.rm = TRUE)
 x
})


```

## Principal Component Analysis On Ames Dataset Numeric Values

### Dataset can be dimensionally reduced PC1 and PC2 graphed below

```{r}
#| echo: false
#| include: false

#testing different dataframes
train <- read.csv(text = c1)
train2 <- read_csv(c1)

#Alley Column had too many NA values so just got rid of it
train_df <- train_df |>  select(-Alley)


#train_df2 <- na.omit(train_df)


#Perform Principal component analysis on numeric columns
train_pca =  train_df |>
 select(where(is.numeric)) |>
 prcomp(scale = TRUE)

 #train_fit <-  train |>
 #select(where(is.numeric)) |>
 #scale() |>
 #prcomp(retx = TRUE)
 #train_fit
```

```{r}
#| echo: false
#| include: false


summary(train_pca)
```

```{r}
#| echo: false
#| include: false


train_pca |> augment(train_df)
```

```{r}
#| echo: false


train_pca |>
 augment(train_df) |> 
ggplot(aes(.fittedPC1, .fittedPC2,color = Foundation)) +
 geom_point(size = 1.5) +
 theme_minimal(base_size=16) +
 labs(title="PCA Coordinates",
 x = "PC 1", y= "PC 2",subtitle = "Example")
```

## What Variables Are the Principal Components?{.smaller}

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 5

arrow_style <- arrow(
  angle = 20, length = grid::unit(8, "pt"),
  ends = "first", type = "closed"
)
train_pca |> 
  tidy(matrix = "rotation")  |> 
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  )  |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text_repel(aes(label = column), hjust = 1,vjust=1) +
  xlim(-0.5, 0.5) + ylim(-0.25, 0.75) + 
  coord_fixed() +
  theme_minimal(base_size =16)
```

```{r}

```

## How Many Principle Components should we use?

### \~5

```{r}
#| echo: false


tibble(var_explained = (train_pca$sdev)^2/80) |> ggplot(aes(x=1:38,y=var_explained)) + geom_point() +
  labs(title="Variance Explained",x="Principal Component",y="Variance Explained") +
  theme_minimal(base_size = 16)
```

## Variables within Principal Components 1-5 {.smaller}

```{r}
#| echo: false


loadings <- as.data.frame(train_pca$rotation)


squared_loadings <- loadings^2


contribpca_df <- squared_loadings[, 1:5]
contribpca_df$total_contrib <- rowSums(contribpca_df)


contribpca_df <- contribpca_df[order(-contribpca_df$total_contrib), ]


contribpca_df <- contribpca_df |> 
 rownames_to_column(var = "variable")

head(contribpca_df)



```

```{r}

```




```{r}
#| echo: false
#| include: false

#Dim 1
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC1`))  |> 
arrange(desc(total_contrib)) |> 
  slice_head(n=10)

#Dim 2
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC2`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)

#Dim 3
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC3`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)

#Dim 4
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC4`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)

#Dim 5
contribpca_df  |> 
 group_by(variable )  |> 
   summarise(total_contrib = sum(`PC5`))  |> 
arrange(desc(total_contrib))|> 
  slice_head(n=10)
```

## Multi Component Analysis on Categorical Variables {.smaller} 

```{r}
#| echo: false


#Multi Component analysis on categorical variables
train_mca =  train_df |>
 select(!where(is.numeric)) |>
MCA( ncp = 5, graph = FALSE)



fviz_mca_var(train_mca,
axes = c(1, 2),
repel = TRUE, select.var = list(contrib = 20))

```

```{r}
#| echo: false
#| include: false

train_mca |> summary()
```

```{r}
#| echo: false
#| include: false

#train_mca$var$contrib[, 1]


sort(train_mca$var$contrib[, 1], decreasing = TRUE)


```

```{r}
#| echo: false
#| include: false


sort(train_mca$var$contrib[, 2], decreasing = TRUE)

```

```{r}
#| echo: false
#| include: false


sort(train_mca$var$contrib[, 3], decreasing = TRUE)
```

```{r}
#| echo: false
#| include: false


head(train_mca$var$coord)

```

```{r}
#| echo: false
#| include: false
#compute the contribution of variable to the dimension component
head(train_mca$var$contrib)

sort(train_mca$var$contrib[, 1], decreasing = TRUE)

contrib_df <- as.data.frame(train_mca$var$contrib)
contrib_df$category <- rownames(contrib_df)


contrib_df$variable <- sub("_.*", "", contrib_df$category)
sort(contrib_df$variable, decreasing = TRUE)



```

## How many MCA components to use?

### again \~5

```{r}
#| echo: false


plot(train_mca$eig[, 1],
type = "b",
xlab = "Dimension",
ylab = "Eigenvalue",
main = "Scree Plot of Ames MCA Dimensions")
```

## Which categorical variables to use? {.smaller}

```{r}
#| echo: false



contrib_df  |> 
 group_by(variable)  |> 
   summarise(total_contrib = sum(`Dim 1`))  |> 
arrange(desc(total_contrib)) |> 
slice_head(n = 10)


contrib_df  |> 
 group_by(variable)  |> 
   summarise(total_contrib = sum(`Dim 2`))  |> 
arrange(desc(total_contrib))|> 
slice_head(n = 10)

contrib_df  |> 
 group_by(variable)  |> 
   summarise(total_contrib = sum(`Dim 3`))  |> 
arrange(desc(total_contrib))|> 
slice_head(n = 10)

contrib_df  |> 
 group_by(variable)  |> 
   summarise(total_contrib = sum(`Dim 4`))  |> 
arrange(desc(total_contrib))|> 
slice_head(n = 10)

contrib_df  |> 
 group_by(variable)  |> 
   summarise(total_contrib = sum(`Dim 5`))  |> 
arrange(desc(total_contrib))|> 
slice_head(n = 10)
```



## Building a Model

### Using variables from the top 5 PCA and MCA dimensions

```{r}
mod <-  lm(SalePrice ~ Foundation+ ExterQual+ BsmtQual+KitchenQual+Exterior1st+Exterior2nd+ GarageFinish+ HeatingQC+BsmtFinType1+MasVnrType+ BsmtCond+ BsmtFinType2+ BsmtExposure+ GarageCond+ GarageFinish+ GarageQual+ GarageType+ KitchenQual+ ExterQual+ RoofStyle+ LandSlope+ RoofMatl+ OverallQual+ GrLivArea+GarageCars+GarageArea+TotalBsmtSF+FullBath+X1stFlrSF+TotRmsAbvGrd+ YearBuilt+ X2ndFlrSF+ BsmtFinSF1+ BsmtFullBath+ HalfBath+KitchenAbvGr+GarageYrBlt+ LotArea+ LotFrontage+ YearRemodAdd+ MSSubClass+ Fireplaces+ EnclosedPorch+ BsmtUnfSF+ WoodDeckSF+OverallCond+ BsmtHalfBath+ScreenPorch, data= train_df)
 mod
```

```{r}
#| echo: false
#| include: false


mod |> summary()
mod |> str()
mod |> tidy(conf.int = TRUE)

tidy_ames = tidy(mod)
tidy_ames |>
slice_max(order_by = estimate, n = 20) |> ggplot(mapping = aes(x = reorder(term,estimate),
 y = estimate)) +
 geom_point() +
   scale_y_log10()+
 coord_flip() +
 theme_minimal(base_size=12) +
 labs(title="Display of Coordinates")
```

## Components of the Model sorted by p values

```{r}
#| echo: false
#| include: false

mod |> tidy(conf.int = TRUE) |> slice_head(n=20)

```

```{r}
#| echo: false


mod_summary <- summary(mod)

# Extract coefficients and convert to data frame
coef_df <- as.data.frame(mod_summary$coefficients)

# Add row names as a column for variable names
coef_df$term <- rownames(coef_df)

# Sort by p-value (column 4)
coef_df_sorted <- coef_df[order(coef_df$`Pr(>|t|)`), ]

# View the top variables by significance
head(coef_df_sorted, 20)

```

## Model Sales Price Prediction. Overall Quality and How New?

```{r}
#| echo: false

mod |> plot_predictions(condition = c("OverallQual","YearBuilt")) +
 theme_minimal(base_size = 16) +
 labs(title = "Better Quality, Newer is Better",
 x="Overall Quality",
 y="Sales Price")
```

## A Better Basement Makes a Difference

```{r}
#| echo: false
#| include: false

mod |> plot_predictions(condition = c("OverallQual","BsmtQual")) +
 theme_minimal(base_size = 16) +
 labs(title = "Excellent Basement Quality",
 x="Overall Quality",
 y="Sales Price")
```

```{r}

```

```{r}
#| echo: false


plot_predictions(
    mod,conf_level = .99,
    condition = list(
        "OverallQual",
        "BsmtQual"), points = .5, rug = TRUE) +
  theme_minimal(base_size=16) +
  labs(title = "Basement Quality",
       subtitle = "", x = "Overall Quality",y="Sales Price")
```

```{r}

```

## Size of Basement Matters too

```{r}
mod |> plot_predictions(condition = c("TotalBsmtSF","YearBuilt")) +
 theme_minimal(base_size = 16) +
 labs(title = "Bigger Basement Better Price",
 x="Basement Square Feet",
 y="Sales Price")
```

## So Does the Kitchen

```{r}
#| echo: false
#| include: false


mod |> plot_predictions(condition = c("YearBuilt","KitchenQual")) +
 theme_minimal(base_size = 16) +
 labs(title = "Kitchen Quality and Year",
 x="Year Built",
 y="Sales Price")
```

```{r}
#| echo: false


plot_predictions(
    mod,conf_level = .99,
    condition = list(
        "YearBuilt",
        "KitchenQual"), points = .5, rug = TRUE) +
  theme_minimal(base_size=16) +
  labs(title = "Newer House with Better Kitchen",
       subtitle = "", x = "Year Built",y="Sales Price")
```

## Lot Size and Exterior Quality

```{r}
#| echo: false


#ExterQual
#LotArea

plot_predictions(
    mod,conf_level = .99,
    condition = list(
        "LotArea",
        "ExterQual"), points = .5, rug = TRUE) +
  scale_x_log10()+
  theme_minimal(base_size=16) +
  labs(title = "Bigger Lot Better Exterior",
       subtitle = "", x = "Lot Area",y="Sales Price")

```

```{r}
#| echo: false
#| include: false

plot_predictions(
    mod,conf_level = .99,
    condition = list(
        "TotalBsmtSF",
        "YearBuilt"), points = .5, rug = TRUE) +
  scale_x_log10()+
  theme_minimal(base_size=16) +
  labs(title = "Bigger Basement Better Price",
       subtitle = "", x = "Lot Area",y="Sales Price")


```


