---
title: "Has the FED been able to fulfill the mandate given to it by Congress?"
author: "Keith DeNivo"
format: pdf
editor: visual
---

## Feds goals

-   Keep price stability (low Inflation rate)

-   Keep Unemployment Low

-   Use interest rates to control the inflation and unemployment

```{r}
#| echo: false
#| include: false
library(readxl)
library(httr)
library(devtools)
library(RCurl)
library(plyr)
library(dplyr)
library(tidyverse)
library(DescTools)
library(ggpubr)
library(openintro)
library(readr)
library(rvest)
library(fpp3)
library(ggplot2)
library(tsibble)
library(feasts)
library(openxlsx)
library(latex2exp)
library(seasonal)
library(seasonalview)
library(fable)
library(rio)
library(urca)
library(zoo)
library(lubridate)
library(writexl)
library(janitor)

```

```{r}
#| echo: false
#| include: false
c1 <- "https://github.com/division-zero/Data-608/raw/refs/heads/main/Story%202/CPI-USeriesReport-20250430212750_4e0f48.xlsx"
#read in the raw file
temp <- tempfile(fileext = ".xlsx")
#download the file
GET(c1, write_disk(temp))

# Load the file into a dataframe
CPI_df <- read_excel(temp)

# temp file removed
unlink(temp)

head(CPI_df)



c3 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%202/FEDFUNDS.csv")
#read in the raw file
fedfunds_df <- data.frame(read.csv(text = c3 ))
#put the csv into a dataframe
head(fedfunds_df)

c2 <- "https://github.com/division-zero/Data-608/raw/refs/heads/main/Story%202/Unemployment%20rateSeriesReport-20250430212633_850590.xlsx"
#read in the raw file
temp <- tempfile(fileext = ".xlsx")
#download the file
GET(c2, write_disk(temp))

# Load the file into a dataframe
unemploy_df <- read_excel(temp)

# temp file removed
unlink(temp)

head(unemploy_df)

```

```{r}
#| echo: false
#| include: false
#| 
#time to clean

CPI_df <- CPI_df |>  slice(-(1:10)) #remove the header rows

unemploy_df <- unemploy_df |>  slice(-(1:10)) #remove the header rows
head(CPI_df)
head(unemploy_df)

```

```{r}
#| echo: false
#| include: false
# set column names to be the first row
colnames(unemploy_df) <- unemploy_df[1, ]
colnames(CPI_df) <- CPI_df[1, ]

# remove the first row
unemploy_df <- unemploy_df[-1, ]
CPI_df <- CPI_df[-1, ]
```

```{r}
#| echo: false
#| include: false
head(CPI_df)
head(unemploy_df)

```

```{r}
#| echo: false
#| include: false
#change all the named months to their respective numbered month
unemploy_df <- unemploy_df |> 
rename(`1` = Jan, `2` = Feb, `3` = Mar, `4` = Apr, `5` = May, `6` = Jun, `7` = Jul, `8` = Aug, `9` = Sep, `10` = Oct, `11` = Nov, `12` = Dec)

#make a column with the numeric date
unemploy_df2 <- unemploy_df |> 
 pivot_longer(cols = -Year, names_to = "month", values_to = "unemployment") |> mutate(date = paste(Year, month, "1", sep = "-")) |> 
select(date, unemployment)

```

```{r}
#| echo: false
#| include: false

#remove the columns with half averages
CPI_df <- CPI_df[, 1:13]
#make a column with the numeric date
CPI_df <- CPI_df |> 
rename(`1` = Jan, `2` = Feb, `3` = Mar, `4` = Apr, `5` = May, `6` = Jun, `7` = Jul, `8` = Aug, `9` = Sep, `10` = Oct, `11` = Nov, `12` = Dec)

#make a column with the numeric date
CPI_df2 <- CPI_df |> 
 pivot_longer(cols = -Year, names_to = "month", values_to = "CPI") |> mutate(date = paste(Year, month, "1", sep = "-")) |> 
select(date, CPI)

colnames(fedfunds_df) <- c("date","FEDFUNDS")

```

```{r}
#| echo: false
#| include: false

#convert the date column into a date data type
CPI_df3 <- CPI_df2 |> 
mutate(date = gsub("\\.0", "", date), #remove the ".0" in 1999.0
date = as.Date(date, format = "%Y-%m-%d"))

#convert the date column into a date data type
unemploy_df3 <- unemploy_df2 |> 
mutate(date = gsub("\\.0", "", date),
date = as.Date(date, format = "%Y-%m-%d"))

#convert the date column into a date data type
fedfunds_df3 <- fedfunds_df |> 
 mutate( date = as.Date(date, format = "%Y-%m-%d"))

#join all data files by date for easier reference in plotting
econ_df <- fedfunds_df3 |> 
 full_join(CPI_df3, by = "date") |> 
full_join(unemploy_df3, by = "date")
econ_df2 <- econ_df |>  drop_na()



```

```{r}
#| echo: false
#| include: false


econ_df2 <- econ_df2 %>%
 mutate(CPI = as.numeric(CPI))


econ_df2 <- econ_df2 |> 
arrange(date) |> mutate(date_numeric=as.numeric(date), CPI_derivative=(CPI-lag(CPI))/(date_numeric-lag(date_numeric))) |> select(-date_numeric)
```

```{r}
#| echo: false
#| include: false


#plotting the raw data in merged dataframe

#compute the derivative of CPI to see if that may provide a better insight
econ_df2 <- econ_df2 |>  mutate(CPI = as.numeric(CPI))
econ_df2 <- econ_df2 |>  mutate(unemployment = as.numeric(unemployment))


fedplot <- econ_df2 |> ggplot(aes(x=date,y=FEDFUNDS))+
  geom_line()
fedplot

unemployplot <- econ_df2 |> ggplot(aes(x = date, y = unemployment))+
  geom_line()
unemployplot

CPIplot <- econ_df2 |> ggplot(aes(x=date,y=CPI))+
  geom_line() 

CPIplot

CPIdplot <- econ_df2 |> ggplot(aes(x=date,y=CPI_derivative))+
  geom_line() 

CPIdplot
```

## Federal Fund Rate

```{r}
#| echo: false
fedplot <- econ_df2 |> ggplot(aes(x=date,y=FEDFUNDS, color = 'red'))+
  geom_line()+
  labs(title = 'Federal Fund Rate. 
We will explore the relationship between itself,
CPI, and Unemployment')+
  ylab('Federal Fund Rate')+
  guides(color = FALSE)+
  theme_minimal()

fedplot


```

## Unemployment %

```{r}
#| echo: false
unemployplot <- econ_df2 |> ggplot(aes(x = date, y = unemployment, color = 'orange'))+
  geom_line() +
  labs(y = 'Unemployment %', title = 'Umemployment Fluctuation') +
  guides(color = FALSE)+
  theme_minimal()
unemployplot


```

## Consumer Price Index

```{r}
#| echo: false


CPIplot <- econ_df2 |> ggplot(aes(x=date,y=CPI))+
  geom_line(color = 'red') +
  labs(title = 'Consumer Price Index Keeps Increasing at different rates')+
  theme_minimal()

CPIplot
```

```{r}
#| echo: false

fedemplot <- econ_df2 |> 
ggplot(aes(x = date)) +
 geom_line(aes(y = FEDFUNDS, color = "Federal Funds Rate")) +
 geom_line(aes(y = unemployment, color = "Unemployment%")) +
 geom_ribbon(aes(ymin = pmin(FEDFUNDS, unemployment), ymax = pmax(FEDFUNDS, unemployment)), alpha = 0.1) +
 labs(color = "Legend",y="%",title = "Federal Fund Rate decreases when Unemployment % increases. 
During periods when employment % is decreasing 
the Federal Fund Rate may increase.") +
 scale_color_manual(values = c("Federal Funds Rate" = "blue", "Unemployment%" = "orange")) +
 theme_minimal()


fedemplot


```

Ideally the area between is minimized near zero %

.

```{r}
#| echo: false
#| include: false
#exploring how the difference between the federal fund rate and unemployment %
econ_df2 <- econ_df2 |> mutate( fed_difference=(unemployment-FEDFUNDS))

```

```{r}
#| echo: false
#| include: false

diffplot <- econ_df2 |> ggplot(aes(x=date,y=fed_difference))+
  geom_line() 
diffplot

#plot basically just looks like the unemployment % plot not very useful
```

## Federal Fund rate is low when Unemployment is High

```{r}
#| echo: false
#exploring the dependence on fed fund rate
fedunplot <- econ_df2 |> ggplot(aes(x=FEDFUNDS,y=unemployment))+
  geom_point(color= 'darkgreen') +
  theme_minimal()+
  labs(title = 'Umeployment % vs Federal Fund Rate', y = "Unemployment %", x = "Federal Fund Rate")
fedunplot


```

```{r}
#| echo: false
#| include: false

#looking at the derivative of CPI with the federal fund rate
fedcpiderplot <- econ_df2 |> ggplot(aes(x=date))+
  geom_line(aes(y=FEDFUNDS))+
  geom_line(aes(y=CPI_derivative))
fedcpiderplot

scale_factor = 50
fedcpiderplot <- econ_df2 |> 
 ggplot(aes(x = date)) +
 geom_line(aes(y = FEDFUNDS, color = "FEDFUNDS")) +
 geom_line(aes(y = CPI_derivative * scale_factor, color = "CPI_derivative")) +
 scale_y_continuous(
 name = "FEDFUNDS",
 sec.axis = sec_axis(~ . / scale_factor, name = "CPI_derivative")) +
 labs(color = "Legend") +
scale_color_manual(values = c("FEDFUNDS" = "blue", "CPI_derivative" = "red")) +
 theme_minimal()

fedcpiderplot


```

## CPI and Federal Fund Rate

```{r}
#| echo: false
scale_factor = 40
fedcpiplot <- econ_df2 |> 
 ggplot(aes(x = date)) +
  geom_line(aes(y = CPI , color = "CPI")) +
 geom_line(aes(y = FEDFUNDS * scale_factor, color = "Federal Fund Rate")) +
  labs(title = 'At first, The Federal Fund Rate does not depend on CPI')+
 scale_y_continuous(
 name = "CPI",
 sec.axis = sec_axis(~ . / scale_factor, name = "Federal Fund Rate")) +
 labs(color = "Legend") +
scale_color_manual(values = c("Federal Fund Rate" = "blue", "CPI" = "red")) +
 theme_minimal()

fedcpiplot
```

```{r}
#| echo: false
#| include: false

#looking at the biweekly rate Change in CPI
days <- 14
econ_df2 <- econ_df2 |> 
 mutate(
 date_numeric = as.numeric(date),
CPI_14derivative = rollapply(CPI, width = days, FUN = function(x) (x[days] - x[1]) / (date_numeric[days] - date_numeric[1]), by = 1, fill = NA, align = "right")
 ) |> 
 select(-date_numeric)
diff14plot <- econ_df2 |> ggplot(aes(x=date,y=CPI_14derivative))+
  geom_line() 
diff14plot


```

## CPI change rate and Federal Fund Rate

```{r}
#| echo: false
scale_factor = 100
fedcpi14plot <- econ_df2 |> 
 ggplot(aes(x = date)) +
 geom_line(aes(y = FEDFUNDS, color = "Federal Fund Rate")) +
  labs(y='Federal Fund Rate', title = 'CPI will react to the Federal Fund Rate.
The federal Fund Rate will increase as a reaction to 
an increase to the CPI rate')+
 geom_line(aes(y = CPI_14derivative * scale_factor, color = "Change in CPI")) +
 scale_y_continuous(
 name = "Federal Fund Rate",
 sec.axis = sec_axis(~ . / scale_factor, name = "Change in CPI")) +
 labs(color = "Legend") +
scale_color_manual(values = c("Federal Fund Rate" = "blue", "Change in CPI" = "red")) +
 theme_minimal()
fedcpi14plot
```

The increases in the federal fund rate as a reaction to the increasing CPI rate are more apparent after 2015.

The federal fund rate does react to the % unemployment and CPI.
