---
title: "Data Story 4: Airbnb Revenue Analysis"
author: "Keith DeNivo"
format: revealjs
editor: visual
---

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(geojsonsf)
library(ggplot2)
library(sf)
library(ggokabeito)
library(knitr)
library(RCurl)
library(kableExtra)
```

```{r}
#| echo: false
#| include: false

#c1 <- getURL("https://github.com/division-zero/Data-608/raw/refs/heads/main/Story%204/neighbourhoods.geojson")




```

```{r}
#| echo: false
#| include: false

c1 <- getURL("https://raw.githubusercontent.com/division-zero/Data-608/refs/heads/main/Story%204/nyc_airbnb_listings%20(1).csv")
#read in the raw file
bnb_df <- data.frame(read.csv(text = c1 ))
#put the csv into a dataframe

head(bnb_df)
```

```{r}
#| echo: false
#| include: false

nyc_geo <- geojson_sf(
"C:/Users/keith/OneDrive - City University of New York/CUNY SPS MS DATA SCIENCE/Data 608/data 608 assignments/Data-608/Story 4/neighbourhoods.geojson")


```

## The Boroughs of New York City

```{r}
#| echo: false
nyc_geo |> ggplot(aes(fill= neighbourhood_group)) +
geom_sf() +
guides(fill = guide_none())+
  scale_fill_okabe_ito()


```

## Price in Each Borough

```{r}
#| echo: false


ggplot(bnb_df, aes(x = borough, y = revenue)) +
geom_violin(fill = "skyblue", color = "black") +
  scale_y_continuous(limits = c(0, 500)) +
labs(title = "price by Borough
     Distribution is Widest in Manhattan",
x = "Borough",
y = "Revenue") +
theme_minimal()+
  coord_flip()



```

## Revenue in Each Borough

```{r}
#| echo: false


ggplot(bnb_df, aes(x = borough, y = revenue)) +
geom_boxplot(, color = "black") +
scale_y_continuous(limits = c(0, 500)) +
labs(title = "Revenue by Borough
     Manhattan Generates the most Revenue",
x = "Borough",
y = "Revenue") +
theme_minimal()
```

```{r}
#| echo: false
#| include: false

nyc_geo |> ggplot(aes(fill= neighbourhood)) +
geom_sf() +
guides(fill = guide_none())
```

```{r}
#| echo: false
#| include: false

cor_df <- bnb_df |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")




```

```{r}
#| echo: false
#| include: false

strong_corrs <- cor_df |> 
  filter(abs(Correlation) > 0.5, Var1 != Var2) |> 
rowwise() |> 
mutate(pair = paste(sort(c(Var1, Var2)), collapse = "_")) |> 
distinct(pair, .keep_all = TRUE) |> 
select(-pair)

revenue_cor <- cor_df |> filter(Var1=="revenue")

unique(bnb_df$neighbourhood_group)

```

```{r}
#| echo: false
#| include: false

mancor_df <- bnb_df |> filter(neighbourhood_group=="Manhattan") |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

manrevenue_cor <- mancor_df |> filter(Var1=="revenue")
manprice <- mancor_df |> filter(Var1=="price")
```

```{r}
#| echo: false
#| include: false

brookcor_df <- bnb_df |> filter(neighbourhood_group=="Brooklyn") |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

brookrevenue_cor <- brookcor_df |> filter(Var1=="revenue")
brookprice <- brookcor_df |> filter(Var1=="price")
```

```{r}
#| echo: false
#| include: false

queencor_df <- bnb_df |> filter(neighbourhood_group=="Queens") |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

Queenrevenue_cor <- queencor_df |> filter(Var1=="revenue")
Queenprice <- queencor_df |> filter(Var1=="price")
```

```{r}
#| echo: false
#| include: false

Bronxcor_df <- bnb_df |> filter(neighbourhood_group=="Bronx") |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

Bronxrevenue_cor <- Bronxcor_df |> filter(Var1=="revenue")
Bronxprice <- Bronxcor_df |> filter(Var1=="price")
```

```{r}
#| echo: false
#| include: false

sicor_df <- bnb_df |> filter(neighbourhood_group=="Staten Island") |> 
select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

sirevenue_cor <- sicor_df |> filter(Var1=="revenue")
siprice <- sicor_df |> filter(Var1=="price")
```

```{r}
#| echo: false
#| include: false

revenue_neighborhood_df <- bnb_df |> 
group_by(neighborhood) |> 
summarize(median_revenue = median(revenue, na.rm = TRUE)) |> 
ungroup()

print(revenue_neighborhood_df)

borough_revenue_df <- bnb_df |> 
group_by(borough) |> 
summarize(median_revenue = median(revenue, na.rm = TRUE)) |> 
ungroup()

print(borough_revenue_df)



top10_n <- revenue_neighborhood_df |> 
arrange(desc(median_revenue)) |> 
head(10)
top10_n
```

```{r}
#| echo: false
#| include: false

```

```{r}
#| echo: false
#| include: false

revenue_neighborhood_df <- revenue_neighborhood_df |> 
 rename(neighbourhood = neighborhood)
```

```{r}
#| echo: false
#| include: false

nyc_geo_revenue <- nyc_geo |> 
left_join(revenue_neighborhood_df, by = "neighbourhood")

```

## Neighborhoods with Most Revenue

```{r}
#| echo: false


nyc_geo_revenue |> ggplot() +
geom_sf(aes(fill= median_revenue), color = "white") +
  scale_fill_gradient(
    name= "Median Revenue",
    low = "white", high= "darkgreen",
    na.value= "grey90"
  )+
labs(title = "NYC Neighborhoods by Median Revenue") +
 theme_minimal() 



```

## Neighborhood List

```{r}
#| echo: false


revenue_neighborhood_df |> 
arrange(desc(median_revenue)) |> slice(1:20) |>  ggplot(aes(x = reorder(neighbourhood, median_revenue), y=median_revenue, fill= median_revenue))+
   scale_y_continuous(
 name = "Median Revenue"
 )+
  geom_col(aes(fill= median_revenue), color = "white")+
  scale_fill_gradient(
    name= "Median Revenue",
    low = "#c1f2bc", high= "darkgreen",
    na.value= "grey90"
  )+
  coord_flip()+
 labs(x="neighborhood",title="Top 20 neighborhoods by Revenue")+
  theme_minimal()
```

```{r}
#| echo: false
#| include: false


ggplot(bnb_df, aes(x = borough, y = revenue)) +
geom_violin(fill = "skyblue", color = "black") +
  scale_y_continuous(limits = c(0, 500)) +
labs(title = "price by Borough",
x = "Borough",
y = "Revenue") +
theme_minimal()#+
 # coord_flip()

ggplot(bnb_df, aes(x = borough, y = revenue)) +
geom_boxplot(, color = "black") +
scale_y_continuous(limits = c(0, 500)) +
labs(title = "Revenue by Borough",
x = "Borough",
y = "Revenue") +
theme_minimal()
  

```

```{r}
#| echo: false
#| include: false

cccor_df <-  bnb_df |> filter(neighborhood==c("Civic Center")) |> select(where(is.numeric)) |> 
cor(use = "complete.obs") |> 
as.data.frame() |> 
rownames_to_column("Var1") |> 
pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")

ccrevenue_cor <- cccor_df |> filter(Var1=="revenue")
ccprice <- cccor_df |> filter(Var1=="price")
```

## What Factors Influence Revenue in Civic Center?

```{r}
#| echo: false


top15_cc <- ccrevenue_cor |> 
arrange(desc(Correlation)) |> 
head(15)


kable(top15_cc, caption = "Top 15 Correlated Variables with Revenue") |> 
 kable_styling(font_size = 10)


```

## What Can we modify to Improve Revenue

What Lessons can we learn from Civic Center?

```{r}
#| echo: false


#cc_df <- bnb_df |> filter(neighborhood==c("Civic Center", "NoHo"))


cc_df <- bnb_df |> filter(neighborhood==c("Civic Center"))

cc_df |> ggplot(aes(x= beds, y = revenue))+
  geom_point(size=3)+
  scale_fill_okabe_ito()+
  geom_smooth(method = "lm", se = FALSE)+
    labs(title="Increase the Number of Beds")+
  theme_minimal(base_size = 20)

#bnb_df |> filter(borough=="Manhattan") |> ggplot(aes(x= beds, y = revenue))+
#  geom_point()+
 # geom_smooth(method = "lm", se = FALSE)
```

```{r}
#| echo: false
#| include: false

#cc_df <- bnb_df |> filter(neighborhood==c("Civic Center"))

cc_df |> ggplot(aes(x= bedrooms, y = revenue))+
  geom_jitter(size=3)+
  scale_fill_okabe_ito()+
  geom_smooth(method = "lm", se = FALSE)+
  theme_minimal(base_size=20)
```

## Work on Increasing Review Score Rating

```{r}
#| echo: false


cc_df |> ggplot(aes(x= review_scores_rating, y = revenue))+
  geom_point(size=3)+
  scale_fill_okabe_ito()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(title="Review Score")+
  theme_minimal(base_size = 20)
```

## Keep Your Description Accurate

```{r}
#| echo: false


cc_df |> ggplot(aes(x= review_scores_accuracy, y = revenue))+
  geom_point(size=3)+
  scale_fill_okabe_ito()+
  geom_smooth(method = "lm", se = FALSE)+
  theme_minimal(base_size = 20)
```

## Keep the Air BNB clean

```{r}
#| echo: false


cc_df |> ggplot(aes(x= review_scores_cleanliness, y = revenue))+
  geom_point(size=3)+
  scale_fill_okabe_ito()+
  labs(title="Cleanliness")+
  geom_smooth(method = "lm", se = FALSE)+
  theme_minimal(base_size = 20)
```

## Make Sure it is Believed to Be a Good Deal

```{r}
#| echo: false


cc_df |> ggplot(aes(x= review_scores_value, y = revenue))+
  geom_point(size=3)+
  scale_fill_okabe_ito()+
  labs(title="Good Value")+
  geom_smooth(method = "lm", se = FALSE)+
  theme_minimal(base_size = 20)
```

## If Construction is Possible

```{r}
#| echo: false

cc_df |> ggplot(aes(x= bedrooms, y = revenue))+
  geom_jitter(size=3)+
  scale_fill_okabe_ito()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(title="Increase the Number of Bedrooms")+
  theme_minimal(base_size = 20)
```
